---
title: "Movielens Project"
author: "Iván de Luna Aldape"
date: "June 1, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, Libraries}
library(tidyverse)
library(caret)
library(data.table)
library(ggplot2)
```

# Introduction

The Movie Lens project is a model that predicts the user rating based on different variables such as the movie, genre and the user behavior regarding ratings.

```{r EDx Code, message = FALSE, warning = FALSE}
### EDX Code ###
################################################################################
#
# Create edx set, validation set (final hold-out test set)
#
################################################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

# RMSE function
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings-predicted_ratings)^2,na.rm=T))
}
```

# Methods and Analysis

## Data Exploratory Analysis

The number of rows in the edx dataset is 9,000,055 with 6 variables
```{r}
#Rows
nrow(edx)
# Variables
ncol(edx)
```

The variables and their class is and their summary
```{r}
# Class
str(edx)
# Summary
summary(edx)
```


The variable timestamp is of class integer, but it needs to be parsed into datetime format, se we use the lubridate library that will help to do it in an easier way.
```{r}
class(edx$timestamp)
edx$timestamp <- as_datetime(edx$timestamp)
validation$timestamp <- as_datetime(validation$timestamp)
```

Check formatted variables is in new POSIXt format

```{r}
class(edx$timestamp)
class(validation$timestamp)
```

Initially the variables userId, movieId, timestamp may not be meaningful but will be used as identification and aggregation variables in order to expand the analysis and have a better organization of the datasets.
The title variable has a format of "title (year)", so we need to create two columns, for title and for year and apply this to both edx and validation sets.
We can use RegEx or, given that the year component of the title variable is in the last part of the string, as (xxxx), we can subtract it without the parenthesis.
We also need to subtract the title name from title and create the title_name variable
```{r}
edx <- edx %>% mutate(year = as.numeric(str_sub(title, -5,-2)))
validation <- validation %>% mutate(year = as.numeric(str_sub(title, -5,-2)))
edx <- edx %>% mutate(title_name = as.character(str_sub(title, end = -8)))
validation <- validation %>% mutate(title_name = as.character(str_sub(title, end = -8)))
```

The rating date can also be separated for better understanding
```{r}
edx$rating_year <- format(edx$timestamp, "%Y")
validation$rating_year <- format(validation$timestamp, "%Y")
```

Genres are also grouped by different names in the same movie title a movie can have multiple genre classifications, so we need to separate them.
Also, trying to manipulate data with such large strings and classifications can become a problem with memory management.
```{r}
edx <- edx %>% separate_rows(genres, sep="\\|")
validation <- validation %>% separate_rows(genres, sep="\\|")
```

# Analysis and Results

## Analysis
There are 69,878 unique users and 10,676 unique titles, of which the most common rating is 4 with a 2,588,430 count, followed by 3 and 5.
```{r}
edx %>% select(userId) %>% 
  n_distinct()
edx %>% select(title) %>% 
  n_distinct()
edx %>% select(rating) %>% 
  group_by(rating) %>% 
  summarise(n = n()) %>% 
  arrange(-n)
```

It seems that people usually rate movies in whole numbers (i.e. 1,2,3,4,5 vs 0.5,1.5,2.5, etc.)
```{r}
edx %>% select(rating) %>% 
  group_by(rating) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>% 
  ggplot(aes(x = rating, y = n)) + geom_bar(stat = "identity") +
  labs(x = "Rating",
       y = "Rating count",
       title = "Ratings")
```

There seems to by a correlation between rating year and average ratings which started pretty high, 4.33, but it has normalized between 3.4 and 3.6 range.
We need to cast the variable rating_year into numeric for the smoothing process.
```{r}
edx %>% select(rating, rating_year) %>% 
  group_by(rating_year) %>% 
  summarise(mean_rating = mean(rating)) %>% 
  arrange(rating_year)

edx %>% select(rating, rating_year) %>% 
  group_by(rating_year) %>% 
  summarise(mean_rating = mean(rating)) %>% 
  ggplot(aes(x = as.numeric(rating_year), y = mean_rating)) +
  geom_point() +
  geom_smooth()
```

If we do the same with movie ratings, there seems to be a preference for classic movies, which realased around 1931-1962.
```{r}
edx %>% select(rating, year) %>% 
  group_by(year) %>% 
  summarise(mean_rating = mean(rating)) %>% 
  arrange(-mean_rating)
```

And movies in general get lower ratings over time or year of release
```{r}
edx %>% select(rating, year) %>% 
  group_by(year) %>% 
  summarise(mean_rating = mean(rating)) %>% 
  ggplot(aes(x = as.numeric(year), y = mean_rating)) +
  geom_point() +
  geom_smooth()
```

The average rating doesn't change over time, but it can be an effect of 4 being the most common rating. It seems that 1995 had the best ratings overall.

```{r}
edx %>% select(title_name, rating) %>% 
  group_by(title_name) %>% 
  summarise(mean_rating = mean(rating),
            ratings = n()) %>% 
  arrange(-ratings)
```

If we concentrate in the top 5 rated movies, there is a downward trend in time for a decrease in average ratings until  around 2005, then a considerable increase in the avg rating for the next years. Only Toy Story and True Lies have a descending rating in 2009.

```{r}
edx %>% select(title_name, rating_year, rating) %>% 
  group_by(title_name) %>% 
  filter(title_name %in% c("Forrest Gump",
                           "Toy Story",
                           "Jurassic Park",
                           "True Lies",
                           "Aladdin")) %>% 
  group_by(title_name, rating_year) %>% 
  summarise(mean_rating = mean(rating)) %>% 
  ggplot(aes(x = rating_year, y = mean_rating, color = title_name,
             group = title_name)) +
  geom_line()
```

We can also check that there is a correlation between number of ratings and mean rating, which shows a trend into the values between 3.5 and 4, the more ratings the movie has.

```{r}
edx %>% select(title_name, rating) %>% 
  group_by(title_name) %>% 
  summarise(mean_rating = mean(rating), ratings = n()) %>% 
  ggplot(aes(x = ratings, y = mean_rating)) +
  geom_point() +
  geom_smooth()
```

The most amount of reviews by year of release is 1995, followed by 1994 and 1996.
```{r}
edx %>%  select(year) %>% 
  group_by(year) %>% 
  summarise(n = n()) %>% 
  arrange(-n)
```

If we arrange by frequency, 10 years have half the reviews and all of them are within 1993 and 2002.
```{r}
edx %>%  select(year) %>% 
  group_by(year) %>% 
  summarise(n = n()) %>% 
  mutate(freq = n/sum(n)) %>%
  arrange(-n)
```
Arranging by year of release, we can see more clearly the concentration of the movie ratings in time.
```{r}
edx %>%  select(year) %>% 
  group_by(year) %>% 
  summarise(n = n()) %>% 
  mutate(freq = n/sum(n)) %>% 
  ggplot() +
  geom_bar(aes(x = year,
               y = n),
               stat = "identity") +
  labs( x = "Title release year",
        y = "Ratings count")
```

In the case of specific titles Forrest Gump has the most reviews with 124,316 followed by Toy Story and Jurassic Park.
```{r}
edx %>% select(title_name) %>% 
  group_by(title_name) %>% 
  summarise(n = n()) %>% 
  mutate(freq = n/sum(n)) %>% 
  arrange(-n)
```


In the case of genres there is a considerable preference for some such as Drama, Comedy and Action.
```{r}
edx %>% group_by(genres) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>%
  ggplot(aes(x = genres, y = n)) + 
  geom_bar(stat = "identity")+ 
  labs(x = "Genre", y = "Reviews")
```

In the case of users, the number of ratings is around 100
```{r}
edx %>% count(userId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, binwidth=0.2, color="black", show.legend = FALSE) + 
  scale_x_log10() + 
  labs(x = "Number of ratings", y = "Number of users")
```

## Results
First we create a test set to train the models, according to the specifications
```{r}
set.seed(1)
test_index <- createDataPartition(y = edx$rating, 
                                  times = 1,
                                  p = 0.2,
                                  list = FALSE)
train_set <- edx[-test_index,]
test_set <- edx[test_index,]
```

Remove probable coincidences from the other sets so it doesn't repeat values.
```{r}
test_set <- test_set %>% 
  semi_join(train_set, by = "movieId") %>%
  semi_join(train_set, by = "userId")
```
The first Model, Average or Naive Average, gives us an  average movie rating of 3.5269
```{r}
mu <- mean(train_set$rating)
mu
```
The RMSE for mu is not really good, this can be affected by a lot of variables and considerations, such as the trend in time or preference for newer or older movies
```{r}
rmse_mu <- RMSE(test_set$rating, mu)
rmse_mu
```
Create a table to store the results
```{r}
rmse_results <- tibble(method = "Average", RMSE = rmse_mu)
```

Least Squares tries to create a model taking into account some other variables, in this case we will predict our value Y with some values b_variable, which in this case will be defined as the mean difference between the actual rating in the training set vs the mean value.

The second model will thus be defined by the simple or naive average and a movie effect
```{r}
movie_mu <- train_set %>% group_by(movieId) %>% 
  summarize(b_movie = mean(rating - mu))

b_movie <- test_set %>% left_join(movie_mu, by = "movieId") %>% 
  pull(b_movie)

y_movie <- mu + b_movie
```

The RMSE is a little bit better than a simple average with a value of 0.94
```{r}
rmse_movie <- RMSE(test_set$rating, y_movie)
rmse_movie
rmse_results <- rbind(rmse_results, tibble(method = "Movie mu",
                                           RMSE = rmse_movie))
```
Visualizing it in an histogram, most of the values are near 0, with a skewness towards negative values.
```{r}
movie_mu %>% qplot(b_movie,
                   geom= "histogram", 
                   bins = 10,
                   data = .)
```

Now we add the average rating the user gives to the movies it reviews
```{r}
user_mu <- train_set %>%  group_by(userId) %>% 
  left_join(movie_mu, by = "movieId") %>% 
  summarise(b_user = mean(rating - mu - b_movie))

b_user <- test_set %>% left_join(user_mu, 
                                 by = "userId") %>% .$b_user
```

We estamite the ratings with simple average, movie effect and user effect And we get a better RMSE of 0.85749 compared to the previous model.
```{r}
y_user <- mu + b_movie + b_user
rmse_user <- RMSE(test_set$rating, y_user)
rmse_user
rmse_results <- rbind(rmse_results, tibble(method = "User mu",
                                           RMSE = rmse_user))
```

We improved substantialy in the RMSE, we can add now a genre effect in order to try to find a better result.
```{r}
genre_mu <- train_set %>%  group_by(genres) %>%
  left_join(movie_mu, by = "movieId") %>% 
  summarise(b_genre = mean(rating - mu))

b_genre <- test_set %>% left_join(genre_mu, 
                                 by = "genres") %>% .$b_genre
```

We now estimate the model with past effects plus the genre effect and we get a worse RMSE than before (0.8646)
```{r}
y_genre <- mu + b_movie + b_user + b_genre
rmse_genre <- RMSE(test_set$rating, y_genre)
rmse_genre

rmse_results <- rbind(rmse_results, tibble(method = "Genre mu",
                                           RMSE = rmse_genre))

```
So we will use the second model which takes into account a movie and user effect and try it in the validation set

```{r PredictionResults, message = FALSE, warning=FALSE}
validation_prediction <- validation %>%
  left_join(movie_mu, by = "movieId" ) %>% 
  left_join(user_mu , by = "userId") %>%
  mutate(pred = mu + b_movie + b_user) %>%
  pull(pred)
```

We check the RMSE and we get 0.8637, which is a little bit worse than in our test set but not much different.
```{r}
model_valid <- RMSE(validation$rating, validation_prediction)
model_valid

rmse_results <- rbind(rmse_results, tibble(method = "Validation",
                                           RMSE = model_valid))
rmse_results
```



# Conclusion


We can see that the model can be preditive enough with the average movie and user effect. This can be improved with further considerations that may count towards the specific usage of the platform, given that there are some heavy users in terms of ratings, which may imply that the more they rate, the more considerations they take into account for such rating.